---
title: "Day 1: Points"
date: "2024-11-01"
categories: [points, crashapi, tigris]
cache: true
---

## Data

I used the [crashapi](https://github.com/elipousson/crashapi) package to gather fatal vehicle crashes for New York city for 2019-2020. Before looking at the maps, recall that stay-at-home orders were given in 2020 due to COVID-19. Therefore, the number of fatal crashes were expected to decrease in 2020 compared to 2019.

```{r data}
# devtools::install_github("elipousson/crashapi")
library(crashapi)
library(tidyverse)
library(janitor)
library(tigris)
library(patchwork)
```

There are 5 counties in New York City, so I made a function to download county data using the `crashapi` package:

```{r}
get_ny_county_data <- function(county){
  crashapi::get_fars(
    year = c(2019, 2020),
    state = "New York",
    county = county,
    details = FALSE,
    geometry = TRUE
  )
}

ny_counties <-  c("New York County", "Kings County", 
    "Queens County", "Bronx County",
    "Richmond County")

ny_crashes <- purrr::map(
  ny_counties,
  ~ get_ny_county_data(.x)
) 

ny_crashes <- purrr::set_names(ny_crashes, ny_counties) 
```

Then I created two lists of dataframes for each year:

```{r}
ny_crashes19 <- purrr::map(ny_crashes, ~ .x |>
                             janitor::clean_names() |> 
                             dplyr::filter(case_year == "2019"))
ny_crashes20 <- purrr::map(ny_crashes, ~ .x |> 
                             janitor::clean_names() |> 
                             dplyr::filter(case_year == "2020"))
```

### City Boundary

City boundary was downloaded using the `tigris` package:

```{r}
#| echo: true
ny_places <- places(state = "NY", cb = TRUE, progress_bar = FALSE)
nyc_boundary <- ny_places[ny_places$NAME == "New York", ]
```

## Point Maps

Finally, I was ready to visualize the single-vehicle fatal crashes in different counties. The function below does exactly that for a given year and county:

```{r}
plot_county_crashes <- function(crash_list, num, total_vehs = "1"){
  dataset <- crash_list[[num]]
  county <- names(crash_list[num])
  num_cases <- nrow(dataset)
  year <- unique(dataset$case_year)
  
  ggplot() +
  geom_sf(data = nyc_boundary, fill = NA, color = "grey50") +
  geom_sf(
    data = dataset |> 
      dplyr::filter(totalvehicles == total_vehs),
    aes(color =  totalvehicles),
    alpha = 0.5
  ) +
  labs(title = paste0(county, " (", year, ")"),
       subtitle = paste0(" (# of crashes: ", num_cases, ")")) +
  theme_void() +
  theme(legend.position = "none")
}
```

And the following function plots the same county for both years:

```{r}
create_combined_plot <- function(num,
                                 list1 = ny_crashes19,
                                 list2 = ny_crashes20
                                 ){
  plot_county_crashes(list1, num) | plot_county_crashes(list2, num)
}
```

### Maps

```{r}
create_combined_plot(1)
create_combined_plot(2)
create_combined_plot(4)
create_combined_plot(5)
```

Bronx county had twice the crashes in 2020 compared to 2019!

And something is wrong with the data for Queens county:

```{r}
create_combined_plot(3)
```


